SIRV-RL: Reinforcement Learning for Optimized Policy Control during Epidemiological Outbreaks in Emerging Market and Developing Economies

### 2024-01-12 paper_draft/main.tex
google_colab_setup for testing, changing the reward function
comparing rewards

At the peak of an epidemic $R_0$ declines as there are no more susceptible people left in the pool, therefore, $R_{eff}$ comes into play. $R_{eff}$ is defined as the average number of secondary cases arising from an infected case, at a given point in an epidemic - therefore, it takes into account the existing immunity of the system.

@@

\begin{equation}
  \beta_{optimal} = 0.04208479828695971
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.02388356686032017
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.7620817917644795
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 85051490.53250012
  \label{eq:cost_without_lockdown}
\end{equation}
See figure \ref{fig:sir_without_lockdown_IND} to see how the model compares with the real data.

@@

\begin{equation}
  \frac{dS}{dt} = -\beta (1 - (s(t)/100)) \frac{S I}{N}
  \label{eq:S_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = -\beta (1 - (s(t)/100)) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I
  \label{eq:R_with_lockdown}
\end{equation}

@@

However, as observed by the value of $\nu_{optimal}$ which is almost negligible and the overestimation of infected individuals in \ref{fig:SIRV_model_with_lockdown_infections_IND} suggests that $\nu$ might be varying with time, i.e., the flow of people from susceptibility to direct recovery changes with a time dependent vaccination rate. Therefore, using the $\beta_{optimal}$ and $\gamma_{optimal}$ from \ref{eq:beta_optimal_with_lockdown_and_nu} and \ref{eq:gamma_optimal_with_lockdown_and_nu}, we optimize the value of $\nu$ given a sub-interval of $[{start}, {start} + {window}]$ where the ${window} = 20$. Here, ${window}$ is a hyperparameter that can be changed so as to reduce the cost function.

@@

Given that the government is an agent that takes decisions in a deterministic environment defined above, we use reinforcement learning to model the competing costs of the environment. This environment is formally as a Markov decision process, and can be described as follows:
\begin{itemize}
    % \item \url{https://jmlr.org/papers/volume22/20-1364/20-1364.pdf} RL Paper that references \url{https://araffin.github.io/slides/rl-tuto-jnrr19/#/13} RL ppt (for our RL notes).
    \item Set of States $\altmathcal{S}$: The state of the environment are described through the descriptors like the normalized GDP ($(GDP_{predicted} - GDP_{min}) / (GDP_{max} - GDP_{min})$), $R_{eff}$, a list of all the previous actions (in changing the stringency) and the proportion of the population that was susceptible, infected and recovered. The starting states are simply these values at the starting date and no previous actions.
    \item Actions $\altmathcal{A}$: The stringency index variable was analyzed with a sample size of 1034. The mean value was approximately 59.62, with a standard deviation of 22.82. The minimum value was 0, while the maximum value reached 100. And the differences between two consecutive stringencies had a mean of 0.030474, and standard deviation of 2.387331, with the minimum being -14.360000 and maximum 55.560000. Based on this we define the discrete action space. There are 7 actions for the agent, it can keep the stringency index same, reduce/increase by 2.5, reduce/increase by 5, and reduce/increase by 10 given that the stringency index doesn't exceed 100 or go below 0.
    \item Transition dynamics $\altmathcal{T}\left(\mathbf{s}_{t+1} \mid \mathbf{s}_t, \mathbf{a}_t\right)$ that map a state-action pair at time $t$ onto a distribution of states at time $t+1$. This state transition is defined by the SIR model with lockdown and the model of how stringency index affects the GDP.
    \item Immediate reward $\altmathcal{R}\left(\mathbf{s}_t, \mathbf{a}_t, \mathbf{s}_{t+1}\right)$. Here we define a reward strategy however it should be noted that this strategy can be easily changed to prioritize different needs.
    \item Discount Factor $\gamma \in[0,1]$, where lower values place more emphasis on immediate rewards. Here, we choose the default discount factor of 0.99.
\end{itemize}
In general, the policy $\pi$ is a mapping from states to a probability distribution over actions: $\pi: \altmathcal{S} \rightarrow p(\altmathcal{A}=\mathbf{a} \mid \altmathcal{S})$. If the MDP is episodic, i.e., the state is reset after each episode of length $T$, then the sequence of states, actions and rewards in an episode constitutes a trajectory or rollout of the policy. Every rollout of a policy accumulates rewards from the environment, resulting in the return $R=\sum_{t=0}^{T-1} \gamma^t r_{t+1}$. The goal of RL is to find an optimal policy, $\pi^*$, which achieves the maximum expected return from all states.

@@

@@

@@

\subsection*{Results}
% https://www.nature.com/articles/s41598-020-79147-8 (see this to write results)
Using the simple SIR model (\ref{eq:S_without_lockdown}, ~\ref{eq:cost_without_lockdown}) to model the disease dynamics we get:
\begin{figure}[H]
  \centering
  \caption{SIR Model with lockdown for India}
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.5]{images/SIR_model_IND.eps}
    \caption{SIR model with lockdown}
    \label{fig:SIR_model_IND}
  \end{subfigure}
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.50]{images/SIR_model_infections_IND.eps}
    \caption{Infections modelled with SIR model}
    \label{fig:SIR_model_infections_IND}
  \end{subfigure}
\end{figure}

Combining the lockdown dynamics in the SIR model we get this following:
\begin{figure}[H]
  \centering
  \caption{SIR Model with lockdown for India}

@@

  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.50]{images/stringency_varying_with_time_IND.eps}
    \caption{Stringency varying with Time}
    \label{fig:stringency_varying_with_time_IND}
  \end{subfigure}
\end{figure}
\begin{figure}[H]\ContinuedFloat
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.50]{images/SIR_model_with_lockdown_infections_IND.eps}
    \caption{Infections modelled with SIR model with Lockdown}
    \label{fig:SIR_model_with_lockdown_infections_IND}
  \end{subfigure}
\end{figure}

Combining vaccination dynamics with SIR with lockdown model we get the following:
\begin{figure}[H]
  \centering
  \caption{SIRV Model with lockdown for India}

@@

\end{figure}

For time-varying $\nu$ we get the folllowing:
\begin{figure}[H]
  \centering
  \caption{SIRV Model with lockdown and time-varying $\nu$ for India}

@@

Stringency affects the economy of developing countries. Below are scatter plots with pearson correlation, for three countries (India, Mexico, Brazil) which are Emerging Market and Developing Economies~\cite{IMFCovid} from May, 2020 to October, 2022.

@@

We observe that stringency has a negative impact on the normalized Gross domestic product (GDP) of each country. Therefore, in some countries policies made during an epidemic have competing costs. This is not the case for Advanced Economies like (USA, Japan, Canada).

>>>>

At the peak of an epidemic $R_0$ declines as there are no more susceptible people left in the pool, therefore, $R_{eff}$ (effective reproductive number) comes into play. $R_{eff}$ is defined as the average number of secondary cases arising from an infected case, at a given point in an epidemic - therefore, it takes into account the existing immunity of the system.

@@

\begin{equation}
  \beta_{optimal} = 0.04208479828695971
  \label{eq:beta_optimal_without_lockdown}
\end{equation}
\begin{equation}
  \gamma_{optimal} = 0.02388356686032017
  \label{eq:gamma_optimal_without_lockdown}
\end{equation}
\begin{equation}
  R_{0_{optimal}} = \frac{\beta_{optimal}}{\gamma_{optimal}} = 1.7620817917644795
  \label{eq:r0_without_lockdown}
\end{equation}
\begin{equation}
  cost = L_{\delta = 1}(S, \hat{S}) + L_{\delta = 1}(I, \hat{I}) + L_{\delta = 1}(R, \hat{R}) = 85051490.53250012
  \label{eq:cost_without_lockdown}
\end{equation}
See figure \ref{fig:SIR_model_IND} to see how the model compares with the real data.

@@

\begin{equation}
  \frac{dS}{dt} = -\beta (1 - (s(t)/100)) \frac{S I}{N}
  \label{eq:S_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dI}{dt} = \beta (1 - (s(t)/100)) \frac{S I}{N} - \gamma I
  \label{eq:I_with_lockdown}
\end{equation}
\begin{equation}
  \frac{dR}{dt} = \gamma I
  \label{eq:R_with_lockdown}
\end{equation}

@@

However, as observed by the value of $\nu_{optimal}$ from \ref{eq:nu_optimal_with_lockdown_and_nu} which is almost negligible (close to zero) and the overestimation of infected individuals in \ref{fig:SIRV_model_with_lockdown_infections_IND} suggests that $\nu$ might be varying with time. This suggests a time-varying vaccination rate, wherein the transition from susceptibility to direct recovery fluctuates with time. Therefore, using the $\beta_{optimal}$ and $\gamma_{optimal}$ from \ref{eq:beta_optimal_with_lockdown_and_nu} and \ref{eq:gamma_optimal_with_lockdown_and_nu}, we optimize the value of $\nu$ given a sub-interval of $[{start}, {start} + {window}]$ where the ${window} = 20$ \cite{Marinov2022}. Here, ${window}$ is a hyperparameter that can be changed so as to reduce the cost function.

@@

Given that the government is an agent that takes decisions in a deterministic environment defined above, we use reinforcement learning to model the competing costs of the environment. This environment is formally as a Markov decision process, and can be described as follows:
\begin{itemize}
    % \item \url{https://jmlr.org/papers/volume22/20-1364/20-1364.pdf} RL Paper that references \url{https://araffin.github.io/slides/rl-tuto-jnrr19/#/13} RL ppt (for our RL notes).
    \item Set of States $\altmathcal{S}$: The state of the environment are described through the descriptors like the normalized GDP ($(GDP_{predicted} - GDP_{min}) / (GDP_{max} - GDP_{min})$), $R_{eff}$, a list of all the previous actions (in changing the stringency) and the proportion of the population that was susceptible, infected and recovered. The starting states are simply these values at the starting date and no previous actions.
    \item TODO: FIX BELOW ITEM AFTER MODEL RUNS
    \item Actions $\altmathcal{A}$: The stringency index variable was analyzed with a sample size of 1034. The mean value was approximately 59.62, with a standard deviation of 22.82. The minimum value was 0, while the maximum value reached 100. And the differences between two consecutive stringencies had a mean of 0.030474, and standard deviation of 2.387331, with the minimum being -14.360000 and maximum 55.560000. Based on this we define the discrete action space. There are 7 actions for the agent, it can keep the stringency index same, reduce/increase by 2.5, reduce/increase by 5, and reduce/increase by 10 given that the stringency index doesn't exceed 100 or go below 0.
    \item Transition dynamics $\altmathcal{T}\left(\mathbf{s}_{t+1} \mid \mathbf{s}_t, \mathbf{a}_t\right)$ that map a state-action pair at time $t$ onto a distribution of states at time $t+1$. This state transition is defined by the SIR model with lockdown and the model of how stringency index affects the GDP.
    \item Immediate reward $\altmathcal{R}\left(\mathbf{s}_t, \mathbf{a}_t, \mathbf{s}_{t+1}\right)$. Here we define a reward strategy however it should be noted that this strategy can be easily changed to prioritize different needs.
    \item Discount Factor $\gamma \in[0,1]$, where lower values place more emphasis on immediate rewards. Here, we choose the default discount factor of 0.99.
\end{itemize}
In general, the policy $\pi$ is a mapping from states to a probability distribution over actions: $\pi: \altmathcal{S} \rightarrow p(\altmathcal{A}=\mathbf{a} \mid \altmathcal{S})$. If the MDP is episodic, i.e., the state is reset after each episode of length $T$, then the sequence of states, actions and rewards in an episode constitutes a trajectory or rollout of the policy. Every rollout of a policy accumulates rewards from the environment, resulting in the return $R=\sum_{t=0}^{T-1} \gamma^t r_{t+1}$. The goal of RL is to find an optimal policy, $\pi^*$, which achieves the maximum expected return from all states.

@@

The reward function is parameterized to account for key factors influencing decision-making. To incentivize reduction of $R{eff}$ (effective reproductive number) a negative reward is imposed of $-20 * R_{eff}$, but as the $R_{eff}$ is between $[1.9, 1,5]$ we start to positively reward it with a multiple of the GDP ($100 * \textrm{min\_max\_normalized\_GDP}$) and if the $R_{eff}$ is below $1.5$ then $200 * \textrm{min\_max\_normalized\_GDP}$. Furthermore, there's a simple positive reward of $10$ if $R_{eff}$ is below the threshold value ($1.9$) and negative reward of $-10$ otherwise. To reward not changing the stringencies frequently, we reward the absolute different between the previous stringency and the current stringency negatively ($|s(t) - s(t-1)| * -1 * 5$) [TODO: this factor should be reduced back to 2 from 5]. If the proportion of the infected population were to rise above 0.003 the model is punished ($-5000$) and otherwise rewarded ($+20$). It should be realized there can be an infinite number of ways to design the reward function to be more human and upgrade the way a decision is taken given the situation \cite{AWSDeepRacer}. Therefore, this research act as a  framework for promoting the development of more efficient reward strategies for the same.

@@

For data that varies with time we use a simple LSTM architecture.

@@

\subsection*{Results}
% https://www.nature.com/articles/s41598-020-79147-8 (see this to write results)
Using the simple SIR model (\ref{eq:S_without_lockdown}, ~\ref{eq:cost_without_lockdown}) to model the disease dynamics we get:
\begin{figure}[H]
  \centering
  \caption{SIR Model with lockdown for India}
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.5]{images/SIR_model_IND.eps}
    \caption{SIR model with lockdown}
    \label{fig:SIR_model_IND}
  \end{subfigure}
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.50]{images/SIR_model_infections_IND.eps}
    \caption{Infections modelled with SIR model}
    \label{fig:SIR_model_infections_IND}
  \end{subfigure}
\end{figure}
Here, it can be observed that the SIR model accurately fits the susceptible population and recovered population but overestimates the infected population by a significant margin. Although overestimating the infected population may not always be problematic, in our specific case, it can create complications. This is because our research involves rewarding the agent when the proportion of infected individuals falls below a predetermined threshold. Consequently, an overestimation of the infected population could lead to incorrect decision-making and undesirable outcomes.

Combining the lockdown dynamics in the SIR model (\ref{eq:S_with_lockdown}, ~\ref{eq:cost_with_lockdown}) we get the following:
\begin{figure}[H]
  \centering
  \caption{SIR Model with lockdown for India}

@@

  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.50]{images/stringency_varying_with_time_IND.eps}
    \caption{Stringency varying with Time}
    \label{fig:stringency_varying_with_time_IND}
  \end{subfigure}
\end{figure}
\begin{figure}[H]\ContinuedFloat
  \begin{subfigure}[t]{\textwidth}
    \centering
    \includegraphics[scale=0.50]{images/SIR_model_with_lockdown_infections_IND.eps}
    \caption{Infections modelled with SIR model with Lockdown}
    \label{fig:SIR_model_with_lockdown_infections_IND}
  \end{subfigure}
\end{figure}
Here, it can be observed there's an overestimation of infected individuals, but, the two stages of the epidemic are being accounted for. This is what suggests that there might be depletion of infected individuals through vaccination.

Combining vaccination dynamics with SIR with lockdown model (\ref{eq:S_with_lockdown_and_nu}, ~\ref{eq:cost_with_lockdown_and_nu}) we get the following:
\begin{figure}[H]
  \centering
  \caption{SIRV Model with lockdown for India}

@@

\end{figure}
Here, because the value of $\nu$ (\ref{eq:nu_optimal_with_lockdown_and_nu}) is negligible, it makes no change. However, time-varying $\nu$ shall be able to better account for the these dynamics.

For time-varying $\nu$ and SIRV model with lockdown (\ref{eq:S_with_lockdown_and_time_varying_nu}, ~\ref{eq:cost_with_lockdown_and_time_varying_nu}), we get the folllowing:
\begin{figure}[H]
  \centering
  \caption{SIRV Model with lockdown and time-varying $\nu$ for India}

@@

With a time-varying $\nu$ (vaccination rate) and the effect of lockdown, our model is able to account for the infected individuals and reduce the cost in comparison to all the previously formalized models for the data. This shows how interventions and changes in the way people behave in response of a epidemic \cite{Caldwell2021} play a major role in the way the epidemic unfolds.

However, Non-pharmaceutical Interventions (NPIs) come with costs for developing nations. Below are scatter plots with pearson correlation, for three countries (India, Mexico, Brazil) which are Emerging Market and Developing Economies~\cite{IMFCovid} from May, 2020 to October, 2022.

@@

It can be observed that stringency has a negative impact on the normalized Gross domestic product (GDP). Therefore, in some countries policies made during an epidemic have competing costs. This is not the case for Advanced Economies like (USA, Japan, Canada).
